---
title: "NEON Biorepository Data Tutorial"
author: "Kelsey Yule"
date: "2024-07-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Learning Objectives

In this tutorial, we will use NEON small mammals as an example to learn how to: 

1. Explore DarwinCore records downloaded from the NEON Biorepository data portal: https://biorepo.neonscience.org

2. Summarize NEON Biorepository sample availability based on spatial, temporal, and taxonomic factors.

3. Download related data from the main NEON data portal using the neonUtilities R package.

3. Join NEON data to associated sample data.

4. Compare NEON Biorepository specimen data to field collected data.

5. Develop a list of samples based on analysis of sample-associated field data.

## Background information

The NEON Biorepository is located at Arizona State University and serves as the primary repository for NEON samples and specimens collected across all 81 NEON field sites. 

The NEON Biorepository data portal allows users to search and download records associated with NEON samples and specimens, request samples and specimens for research, and publish sample-associated data. The NEON Biorepository data portal is built on open-source Symbiota software. Symbiota is the most frequently used software in North America for managing natural history collections records. Learn more about Symbiota data portals at https://symbiota.org/

Symbiota portals publish sample, specimen, and observation data following Darwin Core Standard developed by Biodiversity Information Standards (TDWG). This data standard is a stable, straightforward, and flexible framework for compiling biodiversity data from varied sources. Learn more at https://dwc.tdwg.org/




## Getting Started

First, we want to clear out our workspace and load the required packages.

```{r set up, message=FALSE}
# clean out workspace

#rm(list = ls()) # OPTIONAL - clear out your environment
#gc()            # Uncomment these lines if desired

# load packages
library(tidyverse)
library(neonUtilities)

# source .r file with my NEON_TOKEN
# source("my_neon_token.R") # OPTIONAL - load NEON token
# See: https://www.neonscience.org/neon-api-tokens-tutorial

```

