---
title: "NEON Biorepository Data Tutorial"
author: "Kelsey Yule"
date: "2024-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Learning Objectives

In this tutorial, we will use NEON small mammals as an example to learn how to: 

1. Explore DarwinCore records downloaded from the NEON Biorepository data portal: https://biorepo.neonscience.org

2. Summarize NEON Biorepository sample availability based on spatial, temporal, and taxonomic factors.

3. Download related data from the main NEON data portal using the neonUtilities R package.

3. Join NEON data to associated sample data.

4. Compare NEON Biorepository specimen data to field collected data.

5. Develop a list of samples based on analysis of sample-associated field data.

## Research questions

1. How do NEON **field** and **Biorepository** measurements of deermice (_Peromyscus_) **body masses** compare?

2. Can we obtain a **suitable sample list** to study 5 sets of paired DNA and fecal samples from male _Peromyscus leucopus_ individuals at each of 10 NEON field sites?

## Background information

The NEON Biorepository is located at Arizona State University and serves as the primary repository for NEON samples and specimens collected across all 81 NEON field sites. 

The NEON Biorepository data portal allows users to search and download records associated with NEON samples and specimens, request samples and specimens for research, and publish sample-associated data. The NEON Biorepository data portal is built on open-source *Symbiota* software. *Symbiota* is the most frequently used software in North America for managing natural history collections records. Learn more about *Symbiota* data portals at https://symbiota.org/

Symbiota portals publish sample, specimen, and observation data following *Darwin Core Standard* developed by Biodiversity Information Standards (TDWG). This data standard is a stable, straightforward, and flexible framework for compiling biodiversity data from varied sources. Learn more at https://dwc.tdwg.org/

## Getting Started

First, we want to clear out our workspace and load the required packages. If you do not have the required packages installed previously do so with the `install.packages()` function.

```{r set up, message=FALSE, warning=FALSE}
# clean out workspace

# rm(list = ls()) # OPTIONAL - clear out your environment

# load packages
library(tidyverse)
library(neonUtilities)
library(curl)

# source .r file with my NEON_TOKEN
# source("my_neon_token.R") # OPTIONAL - load NEON token
# See: https://www.neonscience.org/neon-api-tokens-tutorial

```

## Load and explore NEON Biorepository data

Here, we read in a csv file of occurrence records downloaded from the NEON Biorepository data portal. The full results are located in the Github repository https://github.com/kyule/neon-biorepo-tutorial and represent a search for all _Peromyscus leucopus_ samples archived at the NEON Biorepository as of 2024-07-30.

```{r load data, message=FALSE, warning=FALSE}
biorepo<-read.csv("https://github.com/kyule/neon-biorepo-tutorial/raw/main/Peromyscus_leucopus_20240730_DwCRecords/occurrences.csv")
```

Let's look at what information is included in a DarwinCore occurrence record.

```{r investigate},message=FALSE}
# What variables exist for the records?

names(biorepo)


```
We see that a large number of DarwinCore fields are present in the results that outline the who, what, where, when, and more of each sample. Let's explore the data in some fields of interest.

```{r explorefields},message=FALSE}
# What variables exist for the records?

names(biorepo)


```